% !BIB TS-program = biber
\documentclass[11pt,onehalf]{beavtex}
%\documentclass[12pt,twoside]{book}

\input{thesis-header.tex}

% things to check or replace
% replace your name here.
\newcommand{\TheAuthors}{Joanna Piper Morgan}
\title{Algorithms and Software Engineering Schemes for Time-Dependent Radiation Transport on Heterogeneous Compute Architectures}
\degree{Doctor of Philosophy}
\major{Mechanical Engineering}

\submitdate{June 9$^{th}$, 2025}
\commencementyear{2025}

\author{\TheAuthors{}}
\doctype{Dissertation}
\department{Mechanical, Industrial, and Manufacturing Engineering}
\depttype{School}
\depthead{Head}

\advisor{Kyle E.~Niemeyer}


\abstract
{
Predicting the movement of neutrons through space and time is important when modeling inertial confinement fusion, pulsed neutron sources, and nuclear criticality safety experiments, among other systems.
Practical solutions to the transport equation are typically found via Monte Carlo or deterministic methods, or clever combinations of the two.
Both these families of methods prove computationally challenging, particularly for highly time-dependent systems, as the quantities of interest are functions of seven independent variables: time in space, and velocity.
Modern high-performance computing (HPCs) systems with heterogeneous architectures (CPUs and GPUs) are enabling high-fidelity modeling of these systems.
However, to exploit new architectures efficiently, software portability schemes and numerical methods are required.
In this dissertation I discuss both novel numerical methods and software engineering design for deterministic (S$_N$) and Monte Carlo transport on modern HPCs.
First, I introduce and analyze an alternative space-parallel (non-sweeping) iteration algorithm for 1D multi-group time-dependent transport on GPUs, making comparisons to a more traditional iteration algorithm.
I also present a method for converging the space-parallel algorithms iterations in fewer cycles for highly scattering problems in the diffusive limit.
The space-parallel algorithm operates better on modern GPUs and allows use of vendor supplied libraries instead of manually writing GPU functions.
As time step size is decreased the space-parallel algorithm converges in fewer iterations, that a second moment method can be used to converge the space-parallel algorithm in the scattering limit.
Second, I develop a Python and Numba software engineering scheme that allows for developing numerical methods rapidly and deploying at scale in a new open-source Monte Carlo transport application called MC/DC.
MC/DC performs similarly as traditionally compiled code on up to 256 nodes of GPU and CPU based HPCs far a time-dependent problem.
I also develop several novel hybrid Woodcock-delta surface tracking methods in MC/DC including a hybrid-in-energy scheme and a method of using the track-length estimator while Woodcock-delta tracking.
I compare various novel methods against traditional algorithms on CPUs and GPUs with four time-dependent problems, including a continuous energy four-phase reactor accident problem with continuously moving control rods.
}


\acknowledgements{
% 
It's hard for me to express in any meaningful way, the gratitude I have in my heart for those who have supported me through this journey.
These last few years have been the hardest of my life so far, but at every step of the way there have been mentors, teachers, colleagues, friends, and family who have provided harbor from the tempest.
As Newton said if I have done anything, it is only because I have stood on the shoulder's of giants.
You, all of you, are my giants.

I must first start by thanking my advisor, Dr.~Kyle Niemeyer.
In the last five years I have felt pushed and challenged, but never alone in the wilds of research.
Dr.~Niemeyer has been a constant supporter and continually pushes me to make sure that my work has the broadest impact it can.
He has been my bureaucratic warrior: always knowing what levers to pull to make sure everything runs on time.
He has somehow figured out how to be involved in all the fields of science he is interested in which is something I hope to emulate in my own career.

This dissertation would not have been possible without the constant support of Dr.~Todd Palmer who, beyond serving as my minor professor, has helped me---and my dyslexia---through the world of radiation transport.
When I talk to Dr. Palmer about I feel the excitement of discovery welling up through my heart. 
He demands excellence and impunes me when I inevitably fall short in frazzled dispare.
Beyond all the academic and professional accolades I could shower upon him the highest praise I think I can offer, is that he is kind, and that he is caring.
He has made me feel that I belong in this field and in this world.

I hope that I can continue to call the both of you my mentors and friends for many years to come.

Dr.~Ilham Variansyah also deserves a special mention.
This is not the first dissertation he has been acknowledged in and will not be the last.
I am in awe of his work ethic which is only rivaled by his huge heart and impish grin.
I extend a very special thanks to Dr.~Braxton Cuneo who is probably the smartest person I have every had the privileged of working with.
He can chew down on problems like no one I have ever met, demands that the tools we all use to make the world a better place, and is an incredible teacher.
I would like to thank all members of my committee including Dr.~Sourabh Apte, Dr.~Madicken Munk, Dr.~Joshua Gess, and Dr.~Mike Bailey.
All of you have supported me with not only your time and guidance through this multi-year long process but with warmth in the department and field.


I would like to thank my amazing colleagues in the Center for Exascale Monte Carlo Neutron Transport: Alex Mote, Dr.~Aaron Reynolds, Jordan Northrop, Caleb Shaw, Melek Derman, Ethan Lam\'e, and Vincent Novillino.
Dr.~Kayla Clements and Dr.~Sam Passman deserve specific mention as we have commiserated through the at-times sisyphusian process of grad school with me.
I have learned so much from each of you and it has been an absolute pleasure to work with on some of the most challenging and interesting problems of our day.
We have truly shared in \textit{our} victories and commiserate \textit{our} losses \textit{together}.

I would also like to thank my colleagues in the Niemeyer Research Group: Malik Jordan, Dr.~Anthony Walker, Diba Behnoudfar, Dr.~Jayani Jayasuriya, Ali Martz, and Jacob Adams.
As well as the broader OSU community: Michelle Gee, Jessy Urban, Pula Coto, Lindsey Randell, Rachael Macafee, Marshall Andersen, Mike Branco-Katcher, Sophie Windenbenner, Massimo Larsen, Sophie Brodish, Eric Hudec, and many many others.
Together this is the finest and most intellectually sophisticated group of colleagues a young academic could ask for.

To my co-authors and colleagues in the radiation transport and scientific computing community writ-large including: Dr.~Alex Long, Dr.~Kendra Long, Dr.~Travis Tehran, Dr.~Tim Burke, Dr.~Steven Hamilton, Dr.~Camille Palmer, Dr.~Damon MacDougall, Dr.~Dimity Anistrotov, Dr.~C.~T.~Kelley, and Dr.~Ryan McClarren.
I hope I am lucky enough continue this difficult, fascinating, and valuable work with such a talented, tough, and pleasant group of researchers.

The high performance computing staff at Lawrence Livermore National Laboratory have been ever present in my email in box where as that is where all the numerical experiments in this dissertation where conducted.
I thank my funders in the Dept. of Energy and the National Nuclear Security Administration and for the last time will get to say,
\emph{This work was supported by the Center for Exascale Monte-Carlo Neutron Transport (CEMeNT) a PSAAP-III project funded by the Department of Energy, grant number: DE-NA003967.}
I have gone to Oregon public schools my entire academic life: k thru 22\nd grade now.
In fact I still get in-state tuition at OSU so I would like to thank the tax payer who entirely funded this work and almost the entirety of my education.

% Preivous teachers
It's a long standing joke among my friends that, ``Piper can't read.`` While false, probably, I did have a hard time learning to read in the early parts of grade school.
In fact I have never been gifted, talented, or in anyway special when it comes to academic pursuits beyond my nerdy interests and dorky disposition.
If I have done any of this it is only on account of the tens to hundreds of fantastic educators I have had through out my whole life: elementary middle, high, undergrad, and now grad school.
They have supported me, cared for me, and loved me even when I couldn't do those things for myself.
The kicker is each one of them has not only done this for me but hundreds maybe even thousands of others.
If we did little more as a society than provide a life-long good education for all we would be a lot father to utopia.

%They have supported me, cared for me, and loved me even when I couldn't do those things for myself.
%The kicker is that each one of them have probably done that for hundreds if not thousands of people.
%They deserve the world and have my undying gratitude for the holy work they do.

% More casual friends
I would like to thank all my friends and acquaintances through out my life, weather we are currently now in-touch or have fallen to be more distant as the years have gone by.
I love that we dance in and out of each others lives as if a rapid version of the dance of the continents.
But know that as we keep getting thrown around, all of you have a friend-in me to call at any time.
%Sam Patton, Le Lani Fernandez, 

% Family
The reason I became an engineer is because I wanted to be like my grandfather Evan.
The reason I try to extend as much grace as I can to everyone is because of my grandmother Alice Fay.
The reason I try to have a quick whit and never hold back my cackle when something is funny is because of my grandmother Maureen.
The reason I demand my philosophies to be predicated on truth and justice is because of my brother Brooks.
The reason I love my family 
The reason I stand up for myself against bullies and hypocrites is because of my sister Maison.
The reason I know everything that I am doing is achievable because of my sister Marissa.
The reason I burry myself in a problem, don't give up, and dig till it's done is because of my bother Ethan.
The reason I am always an optimist and search for the best in ever person no matter seemingly how destitute they are is because of my brother Spencer.
Each one of you has taught me more then you could ever know.
The reason I am all of these things and anything at all is because of my father Lee and my mother Mary.
I love you guys.


% Serioud friends
Over the past few years my soul has been shattered and remade atom-by-atom.
It's the single most breathtakingly beautiful, empowering, and joyful experiences of my life thus far that I get to be a transgender woman.
But most streets are 2-way and the process of transitioning has also been the most painful and shattering
At this point in my transition I will never surrender on this ordinate my God has called me to, and I can say that I treat my calling with the dignity and respect it deserves.
But I didn't always.
It has only been through the love and compassion of a few that I have been able to rise live in the resplendent joy of truth and light and dreams.
Ediwn, Rosie, Ava, Joslin, and Tricia---I owe all of you everything I am and I live the rest of life trying to make up for the kindnesses you have shown me.

% Patti

To all those who I have the privileged of fighting with good humor, kindness, love, and grace for meaning in this world:
we are not now that strength which in old days
Moved earth and heaven, that which we are, we are,
One equal temper of heroic hearts,
Made weak by time and fate, but strong in will
To \textit{strive}, to \textit{seek}, to \textit{find}, and \textit{never} to yield. 

%\vspace{2em}

%\textit{Thank you}

}


\contributors{
The research objectives of this dissertation are explored in five presented manuscripts.
The following outlines the contributions to the included manuscripts of co-authors other than the major professor.

Dr.~Ilham Variansyah and Dr.~Todd Palmer assisted with the mathematics and writing of Chapters \ref{chap:therefore_paper},  \ref{chap:smom_paper}, \ref{chap:joss_paper}, \ref{chap:cise_paper}, and \ref{chap:delta_tracking_paper}.
Dr.~Ilham Variansyah also provided the models for the Kobayashi, Dragon, C5G7, and C5CE simulations implemented in Chapters \ref{chap:cise_paper} and \ref{chap:delta_tracking_paper}.
Dr.~Braxton Cuneo assisted with writing as well as running simulations in Chapter \ref{chap:cise_paper}. 
Dr.~Kayla Clements supported algorithms development and bug fixing in Chapter \ref{chap:delta_tracking_paper}.
All MC/DC developers (especially Dr.~Ilham Variansyah and Dr.~Braxton Cuneo) developed most of the components of MC/DC used in chapters \ref{chap:joss_paper}, \ref{chap:cise_paper}, and \ref{chap:delta_tracking_paper}.
}

\dedication{{To my grandfather Evan, my dog Rutabaga, \\and all the people who make life worth living.}}

\begin{document}

\maketitle

\mainmatter

% Font for everything after pretext pages
%\fontfamily{cmr}\selectfont
% cmr --- native
% ppl -- plaintino -- liked
% pbk -- bookman -- did not liked
% bch -- charter -- not a fan
% qpl -- tgpagella-- ok
% qbk -- tgbonum -- nope
% qtl -- tgtermes -- not egergious
% qcs -- tgschola -- bolder then above, no
% ptm -- mathptmx -- again not egergious
% qtm -- tgtermes -- hmmm
% lmr -- lmodern -- probs my 2nd fav so far <- what the template uses!
% lmdh -- lmodern -- oh god thats terrible
% put -- utopia / fourier -- too cambria math


% It is faster to make a four-inch mirror, then a six-inch mirror, than to make a six-inch mirror. - Thomspons Rule's for First time Telescope Makers
% It's like tockling the tail of the dragon. - Richard Feynman
% True patriotism springs from a belief in the dignity of the individual, freedom and equality not only for Americans but for all people on earth. - Elenor Roosevelt
% Freedom is nothing but a chance to be better. – Albert Camus
% The life of the nation is secure only while the nation is honest, truthful and virtuous. – Frederick Douglass
%  We are responsible for the world in which we find ourselves, if only because we are the only sentient force which can change it. - James Baldwin
% A civilization is not destroyed by wicked people; it is not necessary that people be wicked but only that they be spineless.
% I love America more than any other country in the world, and, exactly for this reason, I insist on the right to criticize her perpetually.


% setting inital commands for the manuscript header pages
\newcommand{\TheTitle}{BLANK}
\renewcommand{\TheAuthors}{BLANK}
\newcommand{\TheAddress}{BLANK}

%%%


\input{1_introduction}


\part{Deterministic Transport}
\label{part:determ}

\input{2_deterministic}
\input{3_therefore_paper}
\input{4_smom_paper}


\part{Monte Carlo Transport}
\label{part:mc}

\input{5_monte_carlo}
\input{6_joss_paper}
\input{7_cise_paper}
\input{8_delta_tracking_paper}


\input{9_conclusions}


\newpage



\printbibliography


% \newpage
\appendix
\input{appendix/pubs}
\input{appendix/spec_rads_app}
\newpage
\input{appendix/mcatk_app}
\newpage
\input{appendix/trt_paper}
\newpage
\input{appendix/mcdc_prof}

\end{document}
