\chapter{Complete List of Publications}
\label{chap:listopusb}

During the duration of my Ph.D. studies I have authored fourteen manuscripts of which: on eleven I have been first author, seven have been published in peer-reviewer conference transactions or proceedings, and four have been published in field relevant journals.

\begin{enumerate}
    \item Chapter \ref{chap:therefore_paper}: \textsc{J. P. Morgan}, \textsc{I. Variansyah}, \textsc{T. S. Palmer}, and \textsc{K. E. Niemeyer}. (2025) One-Cell Inversion for Solving Higher-Order Time-Dependent Radiation Transport on GPUs. Accepted \emph{Nuclear Science and Engineering}. \doi{10.1080/00295639.2025.2510004}. \arxiv{2503.00264}.

    \item Chapter \ref{chap:smom_paper}: \textsc{J. P. Morgan},  \textsc{T. S. Palmer}, and \textsc{K. E. Niemeyer}. Efficient Preconditioning for Space-Parallel One Cell Inversions in Slab Geometry using a Second Moment Method. In preparation for submission to \emph{Journal of Computational and Theoretical Transport}.

    \item Chapter \ref{chap:joss_paper}: \textsc{J. P. Morgan}, \textsc{I. Variansyah}, \textsc{S. Pasmann}, \textsc{K. B. Clements}, et. al. (2024) Monte Carlo / Dynamic Code (MC/DC): An accelerated Python package for fully transient neutron transport and rapid methods development. \emph{Journal of Open Source Software}. 9 (96), 6415. \doi{10.21105/joss.06415}.

    \item Chapter \ref{chap:cise_paper}: \textsc{J. P. Morgan}, \textsc{I. Variansyah}, \textsc{B. Cuneo}, \textsc{T. S. Palmer}, and \textsc{K. E. Niemeyer}. (2025) Performant and Portable Monte Carlo Neutron Transport via Numba. \bf{27}(1), 57--65. \emph{Computing in Science and Engineering (IEEE)}. \doi{10.1109/MCSE.2025.3550863}. \arxiv{2409.04668}.

    \item Chapter \ref{chap:delta_tracking_paper}: \textsc{J. P. Morgan}, \textsc{I. Variansyah}, \textsc{K. B. Clements}, and \textsc{K. E. Niemeyer}. Hybrid Woodcock-delta Tracking Schemes Using a Track-Length Estimator. In preparation for submission to \emph{Journal of Computational and Theoretical Transport}.

    \item \textsc{J. P. Morgan}, \textsc{A. Mote}, \textsc{S. Pasmann}, \textsc{ G. Ridley}, \textsc{T. S. Palmer}, \textsc{K. E. Niemeyer}, \textsc{R. G. McClarren}. The Monte Carlo Computational Summit -- October 25 \& 26, 2023 -- Notre Dame, Indiana, USA. \emph{Journal of Computational and Theoretical Transport}.\\
    \doi{10.1080/23324309.2024.2354401}.
    \arxiv{2402.08161}. 

    \item \textsc{J. P. Morgan}, \textsc{B. Cuneo}, \textsc{I. Variansyah}, \textsc{K. E. Niemeyer}. Enabling GPU portability into the Numba-JITed Monte Carlo particle transport code MC/DC. (2025). \emph{Proceeding of the International Conference on Mathematics and Computational Methods Applied to Nuclear Science and Engineering (ANS M\&C 2025)}. Denver, CO, USA. \doi{10.13182/MC25-47142}. \arxiv{2501.05440}.

    \item \textsc{B. Cuneo}, \textsc{J. P. Morgan}, \textsc{I. Variansyah}, \textsc{K. E. Niemeyer}. Comparing the Performance of MC/DC's on-GPU Event-based Processing Methods in Multigroup and Continuous-energy Problems. \emph{Proceeding of the International Conference on Mathematics and Computational Methods Applied to Nuclear Science and Engineering (ANS M\&C 2025)}. Denver, CO, USA. DOI: \doi{10.13182/MC25-47142}.

    \item \textsc{J. P. Morgan}, \textsc{T. S. Palmer}, \textsc{T. S. Palmer}, and \textsc{K. E. Niemeyer}. (2023). “Exploring One-Cell Inversion Method for Transient Transport on GPU.” \emph{Proceeding of the International Conference on Mathematics and Computational Methods Applied to Nuclear Science and Engineering (ANS M\&C 2023)}. Niagara Falls, Ontario, Canada. \arxiv{2305.13555}.

    \item Appendix \ref{chapter:mcatk_paper}: \textsc{J. P. Morgan}, \textsc{T. J. Trahan}, \textsc{T. P. Burke}, {C. J. Josey}, and \textsc{K. E. Niemeyer}. (2023). Hybrid-Delta Tracking on a Structured Mesh in MCATK. \emph{Proceeding of the International Conference on Mathematics and Computational Methods Applied to Nuclear Science and Engineering (ANS M\&C 2023)}. Niagara Falls, Ontario, Canada. \arxiv{2306.07847}.

    \item \textsc{I. Variansyah}, \textsc{J. P. Morgan}, {J. Northrop}, \textsc{K. E. Niemeyer}, and \textsc{R. G. McClarren}. (2023). Development of MC/DC: a performant, scalable, and portable Python-based Monte Carlo neutron transport code. \emph{Proceeding of the International Conference on Mathematics and Computational Methods Applied to Nuclear Science and Engineering (ANS M\&C 2023)}. Niagara Falls, Ontario, Canada. \arxiv{2305.07636}.
    
    \item \textsc{J. P. Morgan}, \textsc{T. S. Palmer}, and \textsc{K. E. Niemeyer}. (2022). “Automatic Hardware Code Generation for Neutron Transport Applications.” \emph{Transactions of the American Nuclear Society}, v126, p. 318–-320. American Nuclear Society, Anaheim, CA. \doi{10.13182/T126-38137}. \doi{10.5281/zenodo.6646813}.

    \item Appendix \ref{chapter:trt_paper}: \textsc{J. P. Morgan}, \textsc{A. Long}, \textsc{K. Long}, and \textsc{K. E. Niemeyer}. (2022). “Novel MC TRT Method: Vectorizable Variance Reduction for Energy Spectra.” \emph{Transactions of the American Nuclear Society}, v126, p. 276-–278. American Nuclear Society, Anaheim, CA. \doi{10.13182/T126-38066}. \doi{10.5281/zenodo.6643659}.

    \item \textsc{M. Derman}, \textsc{J. P. Morgan}, \textsc{K. E. Niemeyer}, \textsc{T. S. Palmer}. Unnamed paper on iterative mode decomposition. In preparation.
\end{enumerate}